---
title: "Tooling up for OHDSI"
subtitle: Biostat 218
author: "Marc A Suchard @ UCLA"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format:
  revealjs :
    output-ext: revealjs.html
    footer: Biostat 218 - UCLA - Observational Health Data Sciences and Informatics (OHDSI)
    logo: figures/logo.png  
  html:
    theme: cosmo
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
bibliography: "../ohdsi.bib"
csl: "apa.csl"
knitr:
  opts_chunk: 
    fig.align: 'center'
    fig.width: 6
    fig.height: 4
    message: FALSE
    cache: false
---

## Introduction

In this lecture we will learn how to:

- Prepare our R environments for the OHDSI toolstack

- Download a large synthetic healthcare database

- Perform a basic SQL query

> A journey of a thousand miles begins with a single step
>
> Laozi, Tao Te Ching

## OHDSI toolstack

- Web interface (Athena, Atlas)

- HADES

  - R packages, like Cyclops
  - Java (and python) backends

Follow the instructions at <https://ohdsi.github.io/Hades/> ("setting up the R environment")

## RStudio and renv {.scrollable}

For convenience 

- Fork <https://github.com/ucla-biostat-218/ucla-biostat-218.git> to your own GitHub account

- Clone the fork on your local computer

```{bash}
#| eval: false
#| echo: true
# Use `bash` to clone the project
git clone https://github.com/[USERNAME]/ucla-biostat-218.git
```

- Open RStudio by double-clicking `ucla-biostat-218.Rproj`

> If you *love* GUIs: RStudio -> File -> New Project... -> Version control

- Use renv to install HADES dependencies

```{r}
#| eval: false
#| echo: true
renv::restore()
```

## Synthetic data source {.scrollable}

- Eumonia -- who? (used primarily for continuous integration testing)
  - SQLite

- 1M patients: [download the example database](https://www.dropbox.com/scl/fi/rpqks4ud5n7kthy4i8mkg/synthetic.zip?rlkey=pye84vhflfhvfg1jsbeg703j5&st=0zcnlvgy&dl=1) . This may take a while.  Unzip it and place `synthetic.duckdb` in the `data` directory.
  - DuckDB

- Check the connections:

```{r}
#| echo: false
library(DatabaseConnector)
```

```{r}
#| eval: false
#| echo: true
library(DatabaseConnector)

# Using Eunomia -- will download with each R session
connection <- connect(Eunomia::getEunomiaConnectionDetails())

querySql(connection = connection,
         sql = "SELECT COUNT(*) AS subjects FROM person;")

disconnect(connection)
```

<br>

```{r}
#| echo: true
# Using DuckDB
absoluteFileName <- file.path(getwd(), "data", "synthetic.duckdb")

connection <- connect(dbms = "duckdb", server = absoluteFileName)

querySql(connection = connection,
         sql = "SELECT COUNT(*) AS subjects FROM person;")

disconnect(connection)
```

- Don't worry!  We just want the software and data installed properly.  Later in this class, we will learn about:
  - Standardization across different database servers
  - Structured query language (SQL) and how to overcome its many variants
  - Observational Medical Outcomes Partnership (OMOP) common data model (CDM)
